Speed and Precision in Range Analysis

### 2012-04-10 doc/Manuscript/SBLP/paper.tex
Fixed email address.
Added newest changes to the paper, including more charts.

\author{Victor Hugo Sperle Campos, Igor Rafael de Assis Costa,\\
Raphael Hernani Rodrigues and Fernando Magno Quint\~{a}o Pereira}

@@

@@

>>>>

\author{Victor Hugo Sperle Campos, Igor Rafael de Assis Costa,\\
Raphael Ernani Rodrigues and Fernando Magno Quint\~{a}o Pereira}

@@

The greatest source of improvement in our implementation is the use of strongly
connected components.
In order to propagate ranges across the constraint graph, we fragment it
into strongly connected components, collapse each of these components into
single nodes, and sort the resulting directed acyclic graph topologically.
We then solve the range analysis problem for each component individually.
Once we have solved a component, we propagate its ranges to the next
components, and repeat the process until we walk over the entire constraint
graph.
It is well-known that this technique is essential to speedup constraint solving
algorithms~\cite[Sec 6.3]{Nielson99}.
In our case, the results are dramatic, mostly in terms of speed, but also in
terms of precision.
Figure~\ref{fig:impactSCC} shows the speed up that we gain by using strong
components.
We show results for the integer programs in the SPEC CPU 2006 benchmark suite.
In some cases, as in \texttt{xalancbmk} the analysis on strong components is
450x faster.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/impactSCC}
\end{center}
\caption{\label{fig:impactSCC}
(Left) Bars give time to run our analysis without building strong components
divided by time to run the analysis on strongly connected components.
(Right) Bars give precision, in bitwidth reduction, that we obtain with strong
components, dividid by the precision that we obtain without them.
}
\end{figure}

The strong components improve the precision of our growth analysis.
As we see in Figure~\ref{fig:impactSCC}, in some cases, as in \texttt{bzip2},
it is 40\% more precise.
The gains in precision happen because, by completely resolving a component,
we are able to propagate constant intervals to the next components, instead
of propagating intervals that can grow in both directions.
An an example, in Figure~\ref{fig:ex_partition_grow_crop} we pass the range
$[0, 99]$ from variable $k$ to the component that contains variable $j$.
Had we run the analysis in the entire constraint graph, by the time we
applied the growth analysis on $j$ we would still find $k$ bound to
$0, +\infty$.

A naive implementation of range analysis would solve it once for each
function.
However, we can gain in precision by performing it inter-procedurally; that
is, by allowing the results found for a function $f$ to flow into other
functions that $f$ calls.
Figure~\ref{fig:intra} illustrates the inter-procedural analysis for the
program seen in Figure~\ref{fig:overall_view}(a).

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/intra}
\end{center}
\caption{\label{fig:intra}
Example where an intra-procedural implementation would lead to imprecise
results.
}
\end{figure}

Figure~\ref{fig:wholeImpact} shows that the whole program analysis increases
moderately the precision of our analysis.
We show results for the five largest programs in three different categories of
benchmarks: SPEC CPU 2006, the Stanford Suite~\footnote{\texttt{http://classes.engineering.wustl.edu/cse465/docs/BCCExamples/stanford.c}} and
Bitwise~\cite{Stephenson00}.
Our initial goal when developing this analysis was to support a bitwidth-aware
register allocator.
Thus, we measure precision by the average number of bits that our
analysis allows us to save per program variables.
It is very important to notice that we do not consider constants in our statistics of precision.
In other words, we only measure bitwidth reduction in variables that a constant
propagation step could not remove.
Our results for the SPEC programs were disappointingly: on the average, the
intra-procedural version of our analysis saves 5.23\% of the total program
bitwidth.
By using the inter-procedural version we can increase this number to 8.89\%.
A manual inspection of the SPEC programs reveal that this result is expected:
they manipulate files, and their texts do no provide enough
explicit constants to power our analysis up.
However, with numerical benchmarks we fare much better.
On the average our inter-procedural algorithm reduces the bitwidth of the
Stanford benchmarks by 36.24\%.
Finally, for Bitwise we obtain a bitwidth reduction of 12.27\%.
However, this average is brought down by two outliers: \texttt{edge\_detect} and
\texttt{sha}, which cannot be reduced.
The bitwise benchmarks were implemented by Stephenson
{\em et al.}~\cite{Stephenson00} to validate their bitwidth analysis.
Our results are on par with those found by the original authors.
The bitwise programs contain only the \texttt{main} function; thus, different
versions of our algorithm find the same results.

Another way to increase the precision of range analysis is by using a
context-sensitive implementation.
Context-sensitiveness allows us to distinguish different calling sites of
the same function.
Figure~\ref{context} shows way the ability to make this distinction is
important for precision.
In Figure~\ref{fig:context}(a) we have two different calls of function
\texttt{foo}.
A usual way to perform a data-flow analysis inter-procedurally is to
create assignments between formal and actual parameters, as we show in
Figure~\ref{fig:context}(b).
However, in this case the multiple assignment of values to parameters
makes the ranges of these parameters very large, whereas in reality they
are not.
A way to circumvent this source of imprecision is via function inlining,
as we show in Figure~\ref{fig:context}(c).
The results that we can derive for the transformed program are much more
precise, as each input parameter is assigned a single value.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/context}
\end{center}
\caption{\label{fig:context}
Example where a context-sensitive implementation improves the results of
range analysis.
}
\end{figure}

\begin{figure}[t!]
\begin{center}
\includegraphics[width=\textwidth]{images/wholeImpact}
\end{center}
\caption{\label{fig:wholeImpact}
The impact of whole program analysis on precision. Each bar gives precision in
\%bitwidth reduciton.}
\end{figure}

@@

If strong components account for the largest gains in speed, the choice of
a suitable program representation are responsible for the largest gains in
precision.
However, here we no longer have a win-win condition: a more expressive
program representation decreases our runtime, because it increases the
size of the target program.
Nevertheless, the slowdown is inexpressive.
We chose to run our analysis in programs in the Extended Static Single
Assignment (e-SSA) form~\cite{Bodik00}.
The alternative, which gives us a faster, albeit imprecise, analysis, is to run
our algorithm in programs in Static Single Assignment (SSA)
form~\cite{Cytron91}.
Any program in e-SSA form has also the SSA core property: any variable name
has at most one definition site.
The contrary is not true: SSA form programs do not have the core e-SSA
property: any use site of a variable that appears in a conditional test
post-dominates its definition.
The program in Figure~\ref{fig:overall_view}(b) is in e-SSA form.
The live ranges of variables $i_1$ and $j_1$ have been split right after the
conditional test via the assertions that created variables $i_t$ and $j_t$.
The e-SSA format serves well analyses that extract information from definition
sites and conditional tests, and propagate this information forwardly.
Examples include, in addition to range analysis, tainted flow
analysis~\cite{Rimsa11} and array bounds checks elimination~\cite{Bodik00}.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{images/impactESSA}
\end{center}
\caption{\label{fig:impactESSA}
(Left) Bars give the time to run analysis on e-SSA form programs divided by
the time to run analysis on SSA form programs.
(Right) Bars give the size of the e-SSA form program, in number of
assembly instructions, divided by the size of the SSA form program.}
\end{figure}

Figure~\ref{fig:impactESSA} compares these two program representations in
terms of runtime.
As we see in Figure~\ref{fig:impactESSA}(Left), the e-SSA form slows down our
analysis.
In some cases, as in \texttt{xalancbmk}, this slowdown increases execution
time by 71\%.
Runtime increases because the e-SSA form programs are larger than the SSA
form programs, as we show in Figure~\ref{fig:impactESSA}(Right).
However, this growth is small: in none of the integer programs in SPEC
CPU 2006 we verified an increase in code size of more than 9\%.
But, if the e-SSA form slowdowns the analysis runtime, its gains in precision
are remarkable, as we show in Figure~\ref{fig:precESSA}.

\begin{figure}[t!]
\begin{center}
\includegraphics[width=1\textwidth]{images/precESSA}
\end{center}
\caption{\label{fig:precESSA}
The impact of the e-SSA transformation on precision for three different
benchmark suites. Bars give the ratio of precision (in bitwidth reduction),
obtained with e-SSA form conversion divided by precision without e-SSA form
conversion.}
\end{figure}
